PAMELA: an annotation-based Java Modeling Framework

### 2021-02-17 implementation.tex
implementation after Antidote

\subsection{Exposed API at design-time}

@@

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed. This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).

@@

Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developped to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpretated (it uses PAMELA framework and defines only interfaces and API methods).

Unsurprisingly, we quantifiate a significant CPU overhead (from x20). This is due to the fact that \texttt{MethodHandler} mechanism is fully interpretated and contains many hooks, and cannot be compared to a fully-compiled and optimized Java dynamic binding. As well, memory footprint is increased from a x5 factor.

These conclusions must be moderated by the fact that this overhead only applies to a very small portion of code (mostly property accessors methods). A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer. Using PAMELA won't increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case). As we will see in next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.

Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc...).

>>>>

\subsection{Exposed API at design time}

@@

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed. This also provides an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).

@@

Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses PAMELA framework and defines only interfaces and API methods).

Unsurprisingly, we quantify a significant CPU overhead (from $\times$20). This is due to the fact that \texttt{MethodHandler} mechanism is fully interpreted and contains many hooks, and cannot be compared to a fully compiled and optimized Java dynamic binding. As well, memory footprint is increased from an 
$\times$5 factor.

These conclusions must be moderated by the fact that this overhead only applies
to a very small portion of the code (mostly property accessors methods). A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer. Using PAMELA won’t increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case). As we will see in the next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.

Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
